---
title: "Data preparation"
format: html
editor: visual
---

To get started with the data analysis, you must first ensure that the raw data is in the right format. This script prepares the raw data into a `phyloseq` object and produces the .RData files used in the next steps.

Download the script `002_data_preparation.R`, or copy the codes below into your own script.

```{r}
#| echo: false
#| output: false
library(downloadthis)
```

```{r}
#| echo: false
download_file(
  path = "scripts/002_data_preparation.R",
  output_name = "002_data_preparation",
  button_label = "Download script",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

::: callout-tip
### Source another script {.callout-tip}

In this workflow, the first script is used to install and load R packages. To make sure that all packages are installed and loaded, one should run that script before every other script. A quick and easy way to do so is to source the script at the beginning of every other script:

```{r}
#| eval: false
# Load packages
source("001_required_packages.R")
```
:::

### Auxiliary file

The auxiliary file contains data about samples, which is used to filter samples in the `phyloseq` object. This file could be, for example, an Excel sheet, a .txt file or a .csv file.

```{r}
#| eval: false
# Load an Excel sheet (.xlsx)


# Load a text file (.txt or .csv)


```

### Taxonomy table

The taxonomy table was generated

### OTU table

### Format the data

### Save as RData

### 
