---
title: "Data preparation"
format: html
editor: visual
---

To get started with the data analysis, you must first ensure that the raw data is in the right format. This script prepares the raw data into a `phyloseq` object and produces the .RData files used in the next steps.

To follow the example presented on this page directly on your computer, download the script `002_data_preparation.R` and the example raw data by clicking the buttons below. Alternatively, you may copy the codes below into your own script, and use your own data.

```{r}
#| echo: false
#| output: false
library(downloadthis)
```

```{r}
#| layout: [[-10,40,-5,35,-10]]
#| echo: false
download_file(
  path = "../scripts/002_data_preparation.R",
  output_name = "002_data_preparation",
  button_label = "Download script",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

download_dir(
  path = "../raw data",
  output_name = "Silva Nova MAW raw data",
  button_label = "Download example raw data",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

### Load auxiliary data {#auxfile}

The auxiliary file contains data about samples, which is used to filter samples in the `phyloseq` object. This file could be, for example, an Excel sheet, a .txt file or a .csv file.

```{r}
#| eval: false
# Load an Excel sheet (.xlsx)
aux <- read.xlsx("aux_data.xlsx")

# Load a text file (.txt or .csv)
aux <- read.csv("aux_data.csv")

```

### Construct a phyloseq object from BLAST results {#phyloseq}

#### Taxonomy table

The taxonomy table was generated

#### OTU table

#### Format the data

### Save as .RData {#save}
