---
title: "Data preparation"
format: html
editor: visual
bibliography: references.bib
---

To get started with the data analysis, you must first ensure that the raw data is in the right format. This script prepares the raw data into a `phyloseq` object and produces the .RData files used in the next steps.

To follow the example presented on this page directly on your computer, download the script `002_data_preparation.R` and the example raw data by clicking the buttons below. Alternatively, you may copy the codes below into your own script, and use your own data.

```{r}
#| echo: false
#| output: false
library(downloadthis)
```

```{r}
#| layout: [[-10,40,-5,35,-10]]
#| echo: false
download_file(
  path = "../scripts/002_data_preparation.R",
  output_name = "002_data_preparation",
  button_label = "Download script",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

download_dir(
  path = "../raw data",
  output_name = "Silva Nova MAW raw data",
  button_label = "Download example raw data",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

### Load auxiliary data {#auxfile}

Auxiliary data files contain information about the samples and other types of environmental data. This example auxiliary data file contains site metadata and soil physicochemical analyses. Environmental data can be used in later analyses, such as db-RDA, while metadata can be used to filter samples in the `phyloseq` object.

```{r}
#| eval: false
# Load an Excel sheet (.xlsx)
aux <- read.xlsx("aux_data.xlsx")

# Load a text file (.txt or .csv)
aux <- read.csv("aux_data.csv")
```

### Construct a phyloseq object from BLAST results {#phyloseq}

The procedure is exactly the same with the other target genes (ITS and CO1).

#### Taxonomy table

Our 16S taxonomy table was generated using BLASTn search against SILVA v138.1 database [@quast2012]. Make sure your taxonomy table looks similar to this one before continuing to the next steps.

```{r}
# Load BLAST results
BLAST_raw <- read.delim("../raw data/16S_BLAST.txt")

head(BLAST_raw) # Note that the column qseqid is duplicated (qseqid.1)
```

::: callout-caution
## Were your samples part of a larger sequencing run containing other samples?

If your samples were mixed with multiple other samples in a larger library, you may find OTUs present in none of your samples. Make sure to remove superflu data before you proceed to the next steps.

```{r}
#| eval: false
# Remove NAs or all OTUs with sum == 0
ntaxa(otu_table_raw) # 70006
otu_table_raw <- prune_taxa(taxa_sums(otu_table_raw) > 0, otu_table_raw) 
ntaxa(otu_table_raw) # 25794 (63% of the OTUs were absent of this subset of samples)
```
:::

#### OTU table

#### Format the data

### Save as .RData {#save}
