---
title: "Install R and required packages"
editor: visual
format: html
---

## R and RStudio

To work in R, you'll need to install R and RStudio. Both programs are free and easy to download. For an easy step by step, follow the procedure described [here](https://rstudio-education.github.io/hopr/starting.html#using-r).

::: callout-tip
### Set working directory

In each of your script, always start with setting the working directory to avoid issues with folder structure and ensure that your RData is saved in the proper location.

```{r}
#| eval: false
setwd("some_folder/my_working_directory")
```
:::

## Required packages

Some packages from the CRAN are required to run the examples on this website.

Download the script `001_required_packages.R`, or copy the codes below into your own script.

```{r}
#| echo: false
#| output: false
library(downloadthis)
```

```{r}
#| echo: false
download_file(
  path = "../scripts/001_required_packages.R",
  output_name = "001_required_packages",
  button_label = "Download script",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

#### Install packages

```{r}
#| eval: false
# Install packages
# installed from BiocManager ---------------------------------------------------

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# decontam
if (!require("decontam", quietly = TRUE))
  BiocManager::install("decontam")

# phyloseq
if (!require("phyloseq", quietly = TRUE))
  BiocManager::install("phyloseq")

# CRAN packages ----------------------------------------------------------------

if (!require("ggplot2", quietly = TRUE))
  install.packages("ggplot2")

if (!require("ggpubr", quietly = TRUE))
  install.packages("ggpubr")

if (!require("vegan", quietly = TRUE))
  install.packages("vegan")

if (!require("openxlsx", quietly = TRUE))
  install.packages("openxlsx")

if (!require("plyr", quietly = TRUE))
  install.packages("plyr")

if (!require("dplyr", quietly = TRUE))
  install.packages("dplyr")
```

#### Load packages

```{r}
#| eval: false
# Load packages
library(decontam)
library(phyloseq)
library(ggplot2)
library(ggpubr)
library(vegan)
library(openxlsx)
library(plyr)
library(dplyr) # always load dplyr last to avoid conflicts between packages
```

#### Homemade functions

Some additional functions used in this workflow are not part of any R package. These functions are listed below. Make sure to add them to your scripts!

```{r}
#| eval: false
# Homemade functions -----------------------------------------------------------

# Remove taxa from phyloseq objects
remove_taxa <- function(badTaxa, physeq){
  allTaxa <- taxa_names(physeq)
  cleanTaxa <- allTaxa[!(allTaxa %in% badTaxa)]
  return(prune_taxa(cleanTaxa, physeq))
}

# Merge 10 best hits
mergeHits <- function(BLAST10, sp = 98.7, ge = 94.5, fa = 86.5, or = 82,
                      cl = 78.5, ph = 75, ki = 70){
  
  BLAST10_merge <- BLAST10 %>%
    # Replace NA with character string "NA"
    mutate(across(phylum:species, function(x) if_else(is.na(x), "NA", x))) %>% 
    # Split into a list of data frames
    group_split(OTU)
  
  # Loop through the list of data frames
  for (i in 1:length(BLAST10_merge)){
    
    # Assign NULL to variables without binding
    best.species <- best.genus <- best.family <- best.order <- 
      best.class <- best.phylum <- NULL
    
    # Frequency of species
    species <- as.data.frame(table(BLAST10_merge[[i]]$species)) %>%
      `colnames<-`(c("species", "freq")) %>% 
      mutate(species = as.character(species))
    
    sp.lim <- sp/100*nrow(BLAST10_merge[[i]])
    
    if(max(species$freq) >= sp.lim){
      best.species <- species$species[[which(species$freq >= sp.lim)]]
    }
    if(isTRUE(best.species != "NA")){
      BLAST10_merge[[i]] <- BLAST10_merge[[i]] %>%
        filter(species == best.species) %>%
        arrange(desc(pident)) %>% .[1,]
    } else {
      
      # Frequency of genus
      genus <- as.data.frame(table(BLAST10_merge[[i]]$genus)) %>%
        `colnames<-`(c("genus", "freq")) %>% 
        mutate(genus = as.character(genus))
      
      ge.lim <- ge/100*nrow(BLAST10_merge[[i]])
      
      if(max(genus$freq) >= ge.lim){
        best.genus <- genus$genus[[which(genus$freq >= ge.lim)]]
      }
      if(isTRUE(best.genus != "NA")){
        BLAST10_merge[[i]] <- BLAST10_merge[[i]] %>%
          filter(genus == best.genus) %>%
          arrange(desc(pident)) %>% .[1,] %>%
          mutate(species = NA)
      } else {
        
        # Frequency of family
        family <- as.data.frame(table(BLAST10_merge[[i]]$family)) %>%
          `colnames<-`(c("family", "freq")) %>% 
          mutate(family = as.character(family))
        
        fa.lim <- fa/100*nrow(BLAST10_merge[[i]])
        
        if(max(family$freq) >= fa.lim){
          best.family <- family$family[[which(family$freq >= fa.lim)]]
        }
        if(isTRUE(best.family != "NA")){
          BLAST10_merge[[i]] <- BLAST10_merge[[i]] %>%
            filter(family == best.family) %>%
            arrange(desc(pident)) %>% .[1,] %>%
            mutate(species = NA, genus = NA)
        } else {
          
          # Frequency of order
          order <- as.data.frame(table(BLAST10_merge[[i]]$order)) %>%
            `colnames<-`(c("order", "freq")) %>% 
            mutate(order = as.character(order))
          
          or.lim <- or/100*nrow(BLAST10_merge[[i]])
          
          if(max(order$freq) >= or.lim){
            best.order <- order$order[[which(order$freq >= or.lim)]]
          }
          if(isTRUE(best.order != "NA")){
            BLAST10_merge[[i]] <- BLAST10_merge[[i]] %>%
              filter(order == best.order) %>%
              arrange(desc(pident)) %>% .[1,] %>%
              mutate(species = NA, genus = NA, family = NA)
          } else {
            
            # Frequency of class
            class <- as.data.frame(table(BLAST10_merge[[i]]$class)) %>%
              `colnames<-`(c("class", "freq")) %>% 
              mutate(class = as.character(class))
            
            cl.lim <- cl/100*nrow(BLAST10_merge[[i]])
            
            if(max(class$freq) >= cl.lim){
              best.class <- class$class[[which(class$freq >= cl.lim)]]
            }
            if(isTRUE(best.class != "NA")){
              BLAST10_merge[[i]] <- BLAST10_merge[[i]] %>%
                filter(class == best.class) %>%
                arrange(desc(pident)) %>% .[1,] %>%
                mutate(species = NA, genus = NA, family = NA, order = NA)
            } else {
              
              # Frequency of phylum
              phylum <- as.data.frame(table(BLAST10_merge[[i]]$phylum)) %>%
                `colnames<-`(c("phylum", "freq")) %>% 
                mutate(phylum = as.character(phylum))
              
              ph.lim <- ph/100*nrow(BLAST10_merge[[i]])
              
              if(max(phylum$freq) >= ph.lim){
                best.phylum <- phylum$phylum[[which(phylum$freq >= ph.lim)]]
              }
              if(isTRUE(best.phylum != "NA")){
                BLAST10_merge[[i]] <- BLAST10_merge[[i]] %>%
                  filter(phylum == best.phylum) %>%
                  arrange(desc(pident)) %>% .[1,] %>%
                  mutate(species = NA, genus = NA, family = NA, order = NA, class = NA)
              } else {
                
                # Frequency of kingdom
                kingdom <- as.data.frame(table(BLAST10_merge[[i]]$kingdom)) %>%
                  `colnames<-`(c("kingdom", "freq")) %>% 
                  mutate(kingdom = as.character(kingdom))
                
                ki.lim <- ki/100*nrow(BLAST10_merge[[i]])
                
                if(max(kingdom$freq) >= ki.lim){
                  best.kingdom <- kingdom$kingdom[[which(kingdom$freq >= ki.lim)]]
                }
                if(isTRUE(best.kingdom != "NA")){
                  BLAST10_merge[[i]] <- BLAST10_merge[[i]] %>%
                    filter(kingdom == best.kingdom) %>%
                    arrange(desc(pident)) %>% .[1,] %>%
                    mutate(species = NA, genus = NA, family = NA, order = NA, 
                           class = NA, phylum = NA)
                }
              }
            }
          }
        }
      }
    }
  }
  
  BLAST10_merge <- BLAST10_merge %>% map_df(~as.data.frame(.x))
  
  return(BLAST10_merge)
}
```
