---
title: "Global filtering"
format: html
editor: visual
---

A global filtering is typically applied early to preserve only relevant taxonomic assignations and remove very low-count OTUs that are disproportionately driven by sequencing/PCR error and index misassignment, thereby reducing noise and improving downstream compositional and multivariate analyses.

This section documents how to filter OTUs based on taxonomic assignations and filter out singletons and rare OTUs based on absolute or relative abundance thresholds in `phyloseq` objects. To follow the example presented on this page directly on your computer, download the script `004_global_filter.R` and the example raw data by clicking the buttons below. Alternatively, you may copy the codes below into your own script, and use your own data.

```{r}
#| echo: false
#| output: false
library(downloadthis)
library(plotly)
library(phyloseq)
library(dplyr)
library(ggpubr)
library(microViz)
load("../../RData/16S/otu_ctrl_clean.RData")
```

```{r}
#| layout: [[-10,35,-5,40,-10]]
#| echo: false
download_file(
  path = "../../scripts/004_global_filter.R",
  output_name = "004_global_filter",
  button_label = "Download script",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

download_file(
  path = "../../RData/16S/otu_ctrl_clean.RData",
  output_name = "otu_ctrl_clean",
  button_label = "Download example RData",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

## Taxonomy filtering

This first simple step serves to preserve only the relevant OTUs that were targeted with each primer sets: i.e., Bacteria for 16S, Fungi for ITS and Metazoa for CO1.

Note that the 16S primers amplifying the V4-V5 regions are also suitable for Archaea. The choice to include them or not in the following filtering depends on the hypotheses of your study. Similarly, primers targeting the internal transcribed spacer (ITS1 and ITS2) regions, while commonly used as the universal barcoding marker for fungi, often co-amplify DNA from a wide variety of other Eukaryotes due to the conserved nature of the flanking ribosomal RNA genes (18S, 5.8S, and 28S). For example, DNA from plants and protists is often amplified using these primers. It is recommended to filter them out when the hypotheses of the study are focused on Fungi. Finally, CO1 primers may also amplify Prokaryotes or other Eukaryotes, but these are not specificly targeted and should be filtered out.

+----------+------------------------------------+-----------------------------------------+-----------+-------------+---------------------------+
| Name     | Target                             | Primers (sequence 5’-3’)                | Regions   | Lenght      | Reference                 |
+==========+====================================+=========================================+===========+=============+===========================+
| 16S      | Prokaryotes (Bacteria and Archaea) | 515F-Y (GTGYCAGCMGCCGCGGTAA)\           | V4-V5     | 360-500 bp  | [@parada2015]             |
|          |                                    | 926RB (CCGYCAATTYMTTTRAGTTT)            |           |             |                           |
+----------+------------------------------------+-----------------------------------------+-----------+-------------+---------------------------+
| ITS      | Fungi                              | ITS9MUN (TGTACACACCGCCCGTCG)\           | ITS1-ITS2 | 300-1200 bp | [@tedersoo2016]           |
|          |                                    | ITS4ngsUni (CCTSCSCTTANTDATATGC)        |           |             |                           |
+----------+------------------------------------+-----------------------------------------+-----------+-------------+---------------------------+
| CO1      | Metazoa                            | mlCOIintF (GGWACWGGWTGAACWGTWTAYCCYCC)\ | 320-345\  | \~313 bp    | [@leray2013; @geller2013] |
|          |                                    | jgHCO2198 (TAIACYTCIGGRTGICCRAARAAYCA)  | 658       |             |                           |
+----------+------------------------------------+-----------------------------------------+-----------+-------------+---------------------------+

: Table 1. Primer sets used for this data set

This demonstration is done with the 16S data, and we choose to keep only Bacteria.

```{r}
#| output: false
# Count OTUs before filtering
ntaxa(otu_ctrl_clean) # 25390

# Filter Bacteria
otu_table_tax <- subset_taxa(otu_ctrl_clean, kingdom == "Bacteria")

# Count OTUs after filtering
ntaxa(otu_table_tax) # 24790 (~2% of the OTUs were not Bacteria)
```

## Singletons and rare OTUs

Singletons and other rare OTUs are predominantly composed of PCR, sequencing, and chimera artefacts, and their removal has been shown to have negligible effects on hypothesis-driven community comparisons while substantially reducing noise [@edgar2016; @nikodemova2023; @sgarbi2020; @cao2021]. Retaining them is mainly justified in studies explicitly targeting rare biosphere diversity and requires careful validation [@brown2015]. Consequently, filtering rare features simplifies microbiome data sets without materially altering ecological inference in most comparative analyses.

In R, singletons are typically removed by excluding OTUs with a total count of one across all samples, while rare OTUs are filtered by applying absolute (e.g. total counts \<5 or \<10) or relative abundance thresholds (e.g. \<0.01%) to the OTU table [@edgar2016; @brown2015; @nikodemova2023], commonly using `phyloseq::prune_taxa()` in combination with `taxa_sums()`.

### Removing singletons

Removing singletons is as simple as keeping anything that has more than 1 read throughout the entire data set.

```{r}
#| output: false
# Remove singletons
otu_table_single <- prune_taxa(taxa_sums(otu_table_tax) > 1, otu_table_tax)

# Count OTUs after filtering
ntaxa(otu_table_single) # 24339 (~2% of the remaining OTUs were singletons)
```

### Removing rare OTUs

OTUs are considered rare if their abundance is small compared to other OTUs. They may be present in many samples, but if their total sum across samples is smaller than *k*% of the average sequencing depth (relative abundance), or smaller than *k* reads (absolute abundance) then they are considered rare.

NOTE: When calculating the average sequencing depth, ignore controls and blanks, as they artificially lower the average.

#### Relative abundance

Typically, thresholds between 0.005% and 0.1% relative abundance are found in the literature. For the sake of this demonstration, we've included multiple thresholds between 0.005% and 0.1%, as well as a threshold up to 1% to better illustrate the effects of filtering out rare OTUs on different community analyses.

We define relative thresholds based on the average sequencing depth (assuming that all samples have sufficient sequencing depth).

```{r}
#| output: false
# Calculate the average sequencing depth
seq_depth_avg <- mean(sample_sums(otu_table_tax)) # 148149.9 reads

# Calculate thresholds based on k% of the average sequencing depth
# For the sake of this exercise, multiple thresholds between 0.005% and 1% are
# tested to compare their effects on various community analyses.
thresh005 <- ceiling(seq_depth_avg*0.005/100) # 8 reads
thresh01 <- ceiling(seq_depth_avg*0.01/100) # 15 reads
thresh05 <- ceiling(seq_depth_avg*0.05/100) # 75 reads
thresh1 <- ceiling(seq_depth_avg*0.1/100) # 149 reads
thresh10 <- ceiling(seq_depth_avg*1/100) # 1482 reads

# Remove all OTUs that have a total sum lower than the threshold
otu_table_rare005 <- otu_table_tax %>% prune_taxa(taxa_sums(.) > thresh005, .)
otu_table_rare01 <- otu_table_tax %>% prune_taxa(taxa_sums(.) > thresh01, .)
otu_table_rare05 <- otu_table_tax %>% prune_taxa(taxa_sums(.) > thresh05, .)
otu_table_rare1 <- otu_table_tax %>% prune_taxa(taxa_sums(.) > thresh1, .)
otu_table_rare10 <- otu_table_tax %>% prune_taxa(taxa_sums(.) > thresh10, .)

# Compare the number of OTU in each filtering step
ntaxa(otu_ctrl_clean) #25390 (before filtering)
ntaxa(otu_table_tax) #24790 (~98% are Bacteria)
ntaxa(otu_table_single) #24339 (~2% singletons)
ntaxa(otu_table_rare005) #8548 (~66% of OTUs have less than 8 reads; 0.005%)
ntaxa(otu_table_rare01) #6735 (~73% of OTUs have less than 15 reads; 0.01%)
ntaxa(otu_table_rare05) #3765 (~85% of OTUs have less than 75 reads; 0.05%)
ntaxa(otu_table_rare1) #2913 (~88% of OTUs have less than 149 reads; 0.1%)
ntaxa(otu_table_rare10) #1032 (~96% of OTUs have less than 1482 reads; 1%)
```

#### Absolute abundance

Typically, thresholds between 5-10 reads are found in the literature. In this example, this roughly corresponds to a relative abundance threshold at 0.005% (8 reads). If the average sequencing depth of your data set is larger than the one presented here, it may be relevant to test removing 5 and 10 reads if those absolute thresholds are significantly lower than a relative threshold at 0.005%.

#### Compare thresholds

To visualize the effect of these different thresholds, we must first create a data frame per filtered OTU table that summarizes the sequencing depth per sample and the Observed OTU (richness).

```{r}
#| output: false
# Non-filtered OTU table
seq_depth_raw <- cbind.data.frame(sample_data(otu_table_tax)) %>%
  cbind(estimate_richness(otu_table_tax, measures = "Observed")) %>%
  mutate(LibrarySize = sample_sums(otu_table_tax)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "raw")

# Singletons
seq_depth_single <- cbind.data.frame(sample_data(otu_table_single)) %>%
  cbind(estimate_richness(otu_table_single, measures = "Observed")) %>%
  mutate(LibrarySize = sample_sums(otu_table_single)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "singletons")

# Rare 0.005% (<8 reads)
seq_depth_rare005 <- cbind.data.frame(sample_data(otu_table_rare005)) %>%
  cbind(estimate_richness(otu_table_rare005, measures = "Observed")) %>%
  mutate(LibrarySize = sample_sums(otu_table_rare005)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "rare 0.005%")

# Rare 0.01% (<15 reads)
seq_depth_rare01 <- cbind.data.frame(sample_data(otu_table_rare01)) %>%
  cbind(estimate_richness(otu_table_rare01, measures = "Observed")) %>%
  mutate(LibrarySize = sample_sums(otu_table_rare01)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "rare 0.01%")

# Rare 0.05% (<75 reads)
seq_depth_rare05 <- cbind.data.frame(sample_data(otu_table_rare05)) %>%
  cbind(estimate_richness(otu_table_rare05, measures = "Observed")) %>%
  mutate(LibrarySize = sample_sums(otu_table_rare05)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "rare 0.05%")

# Rare 0.1% (<149 reads)
seq_depth_rare1 <- cbind.data.frame(sample_data(otu_table_rare1)) %>%
  cbind(estimate_richness(otu_table_rare1, measures = "Observed")) %>%
  mutate(LibrarySize = sample_sums(otu_table_rare1)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "rare 0.1%")

# Rare 1% (<1482 reads)
seq_depth_rare10 <- cbind.data.frame(sample_data(otu_table_rare10)) %>%
  cbind(estimate_richness(otu_table_rare10, measures = "Observed")) %>%
  mutate(LibrarySize = sample_sums(otu_table_rare10)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "rare 1%")

# Combine data frames
seq_depth_all <- seq_depth_raw %>% rbind(seq_depth_single) %>%
  rbind(seq_depth_rare005) %>% rbind(seq_depth_rare01) %>%
  rbind(seq_depth_rare05) %>% rbind(seq_depth_rare1) %>%
  rbind(seq_depth_rare10)
```

##### Effect on library size

```{r}
#| eval: false
# Plot the library size of all samples
ggplot(data = seq_depth_all, aes(x = Index, y = LibrarySize, color = seq_depth)) +
  geom_point() + geom_line()
```

```{r}
#| echo: false
plot_ly(data = seq_depth_all, x = ~Index, y = ~LibrarySize, color = ~seq_depth,
        type = 'scatter', mode = 'markers+lines', colors = 'Dark2', 
        text = ~paste("Library Size: ", LibrarySize, 
                      '<br>Sample ID: ', Soil_ID))
```

With this analysis, we can see that the first filtering step (singletons) does not affect LibrarySize, as the points completely overlap (yellow vs brown). Aside from a threshold at 1%, other thresholds barely reduced library size per sample.

::: callout-tip
Try zooming in on a selection of points with low library size or large library size to better see the effects of the different thresholds. For example, the filtering of rare OTUs at a threshold of 0.01% changes very little in the LibrarySize, even though we lost \~74% of OTUs.
:::

##### Effect on OTU richness

To confirm our choice of the best filtering threshold, let's compare the effect of filtering on OTU richness.

```{r}
#| eval: false
# Plot the OTU richness of all samples
ggplot(data = seq_depth_all, aes(x = Index, y = Observed, color = seq_depth)) +
  geom_point() + geom_line()
```

```{r}
#| echo: false
plot_ly(data = seq_depth_all, x = ~Index, y = ~Observed, color = ~seq_depth,
        type = 'scatter', mode = 'markers+lines', colors = 'Dark2', 
        text = ~paste("Library Size: ", LibrarySize, 
                      "<br>Observed OTU: ", Observed, 
                      '<br>Sample ID: ', Soil_ID))
```

This plot shows that these different filtering thresholds return the same pattern of OTU richness as a function of library size.

```{r}
#| echo: false
#| output: false
# Non-filtered OTU table
seq_depth_raw <- cbind.data.frame(sample_data(otu_table_tax)) %>%
  cbind(estimate_richness(otu_table_tax, measures = "Observed")) %>%
  cbind(estimate_richness(otu_table_tax, measures = "Shannon")) %>%
  mutate(LibrarySize = sample_sums(otu_table_tax)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "raw")

# Singletons
seq_depth_single <- cbind.data.frame(sample_data(otu_table_single)) %>%
  cbind(estimate_richness(otu_table_single, measures = "Observed")) %>%
  cbind(estimate_richness(otu_table_single, measures = "Shannon")) %>%
  mutate(LibrarySize = sample_sums(otu_table_single)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "singletons")

# Rare 0.005% (<8 reads)
seq_depth_rare005 <- cbind.data.frame(sample_data(otu_table_rare005)) %>%
  cbind(estimate_richness(otu_table_rare005, measures = "Observed")) %>%
  cbind(estimate_richness(otu_table_rare005, measures = "Shannon")) %>%
  mutate(LibrarySize = sample_sums(otu_table_rare005)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "rare 0.005%")

# Rare 0.01% (<16 reads)
seq_depth_rare01 <- cbind.data.frame(sample_data(otu_table_rare01)) %>%
  cbind(estimate_richness(otu_table_rare01, measures = "Observed")) %>%
  cbind(estimate_richness(otu_table_rare01, measures = "Shannon")) %>%
  mutate(LibrarySize = sample_sums(otu_table_rare01)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "rare 0.01%")

# Rare 0.05% (<76 reads)
seq_depth_rare05 <- cbind.data.frame(sample_data(otu_table_rare05)) %>%
  cbind(estimate_richness(otu_table_rare05, measures = "Observed")) %>%
  cbind(estimate_richness(otu_table_rare05, measures = "Shannon")) %>%
  mutate(LibrarySize = sample_sums(otu_table_rare05)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "rare 0.05%")

# Rare 0.1% (<151 reads)
seq_depth_rare1 <- cbind.data.frame(sample_data(otu_table_rare1)) %>%
  cbind(estimate_richness(otu_table_rare1, measures = "Observed")) %>%
  cbind(estimate_richness(otu_table_rare1, measures = "Shannon")) %>%
  mutate(LibrarySize = sample_sums(otu_table_rare1)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "rare 0.1%")

# Rare 1% (<1508 reads)
seq_depth_rare10 <- cbind.data.frame(sample_data(otu_table_rare10)) %>%
  cbind(estimate_richness(otu_table_rare10, measures = "Observed")) %>%
  cbind(estimate_richness(otu_table_rare10, measures = "Shannon")) %>%
  mutate(LibrarySize = sample_sums(otu_table_rare10)) %>%
  arrange(LibrarySize) %>% mutate(Index = row_number()) %>%
  mutate(seq_depth = "rare 1%")

# Combine data frames
seq_depth_all <- seq_depth_raw %>% rbind(seq_depth_single) %>%
  rbind(seq_depth_rare005) %>% rbind(seq_depth_rare01) %>%
  rbind(seq_depth_rare05) %>% rbind(seq_depth_rare1) %>%
  rbind(seq_depth_rare10)
```

::: callout-tip
Try comparing other alpha diversity indices, such as the Shannon index.

```{r}
#| eval: false
# Add Shannon index to each OTU table as such
seq_depth_raw <- seq_depth_raw %>%
  cbind(estimate_richness(otu_table_tax, measures = "Shannon"))

# Plot the Shannon index of all samples
ggplot(data = seq_depth_all, aes(x = Index, y = Shannon, color = seq_depth)) +
  geom_point() + geom_line()
```

```{r}
#| echo: false
plot_ly(data = seq_depth_all, x = ~Index, y = ~Shannon, color = ~seq_depth,
        type = 'scatter', mode = 'markers+lines', colors = 'Dark2', 
        text = ~paste("Library Size: ", LibrarySize, 
                      "<br>Shannon index: ", Shannon, 
                      '<br>Sample ID: ', Soil_ID))
```

Filtering rare OTUs has no significant effect on the Shannon index. Note also the lack of a clear relationship between the Shannon index and LibrarySize.
:::

##### Effect on beta diversity

To confirm our choice of filtering further, we can compare how the different filtering methods affect beta diversity.

```{r}
# Non-filtered OTU table
beta_div_raw <- otu_table_tax %>%
  tax_transform("identity", rank = "taxa_OTU") %>%
  dist_calc("aitchison") %>%
  ord_calc("PCoA") %>%
  ord_plot(color = "Land_use", auto_caption = NA) +
  stat_ellipse(aes(colour = Land_use))

# Singletons
beta_div_single <- otu_table_single %>%
  tax_transform("identity", rank = "taxa_OTU") %>%
  dist_calc("aitchison") %>%
  ord_calc("PCoA") %>%
  ord_plot(color = "Land_use", auto_caption = NA) +
  stat_ellipse(aes(colour = Land_use))

# Rare 0.005%
beta_div_rare005 <- otu_table_rare005 %>%
  tax_transform("identity", rank = "taxa_OTU") %>%
  dist_calc("aitchison") %>%
  ord_calc("PCoA") %>%
  ord_plot(color = "Land_use", auto_caption = NA) +
  stat_ellipse(aes(colour = Land_use))

# Rare 0.01%
beta_div_rare01 <- otu_table_rare01 %>%
  tax_transform("identity", rank = "taxa_OTU") %>%
  dist_calc("aitchison") %>%
  ord_calc("PCoA") %>%
  ord_plot(color = "Land_use", auto_caption = NA) +
  stat_ellipse(aes(colour = Land_use))

# Rare 0.05%
beta_div_rare05 <- otu_table_rare05 %>%
  tax_transform("identity", rank = "taxa_OTU") %>%
  dist_calc("aitchison") %>%
  ord_calc("PCoA") %>%
  ord_plot(color = "Land_use", auto_caption = NA) +
  stat_ellipse(aes(colour = Land_use))

# Rare 0.1%
beta_div_rare1 <- otu_table_rare1 %>%
  tax_transform("identity", rank = "taxa_OTU") %>%
  dist_calc("aitchison") %>%
  ord_calc("PCoA") %>%
  ord_plot(color = "Land_use", auto_caption = NA) +
  stat_ellipse(aes(colour = Land_use))

# Rare 1%
beta_div_rare10 <- otu_table_rare10 %>%
  tax_transform("identity", rank = "taxa_OTU") %>%
  dist_calc("aitchison") %>%
  ord_calc("PCoA") %>%
  ord_plot(color = "Land_use", auto_caption = NA) +
  stat_ellipse(aes(colour = Land_use))

# Visualize all plots together
ggarrange(beta_div_raw, beta_div_single, beta_div_rare005, beta_div_rare01,
          beta_div_rare05, beta_div_rare1, beta_div_rare10,
          labels = c(
            "Non-filtered (24790 OTUs)", "Singletons (24339 OTUs)",
            "0.005% (<8 reads; 8548 OTUs)", "0.01% (<15 reads; 6735 OTUs)",
            "0.05% (<75 reads; 3765 OTUs)", "0.1% (<149 reads; 2913 OTUs)",
            "1% (<1482 reads; 1032 OTUs)"),
          ncol = 3, nrow = 3, common.legend = TRUE,
          hjust = 0, label.x = 0.2, font.label = list(size = 10))
```

With this analysis, we can see that removing rare OTUs reduces the variance within treatment a little bit, but does not affect overall patterns. Overall, all three plots showed that rare OTUs were mostly found in samples with large library size, and filtering out rare OTUs does not affect patterns. Only a threshold as high as 1% considerably affected patterns.

#### Choosing the best threshold

Ultimately, the choice of the "best" threshold is completely arbitrary. Various thresholds are found in the literature and no one is more justified than another. As a reminder: singletons and other rare OTUs have been previously found to be predominantly composed of artefacts, thus, their removal has been shown to have negligible effects on hypothesis-driven community comparisons while substantially reducing noise [@edgar2016; @nikodemova2023; @sgarbi2020; @cao2021]. Retaining them is only justified in studies explicitly targeting rare biosphere diversity and requires careful validation [@brown2015].

As seen in this demonstration, all thresholds below 1% had little observable effect on library size, alpha diversity and beta diversity. For the sake of being more conservative, let's select the lowest relative abundance threshold (0.005%), which also roughly corresponded to a conservative absolute threshold (5-10 reads).

```{r}
# Select the best filtering threshold (0.005%)
# ~66% of OTUs have less than 8 reads
otu_table_rare <- otu_table_rare005
```

### Clean and save OTU table

```{r}
#| eval: false
# OTU table with singletons filtered
save(otu_table_single, file = "RData/16S/otu_single_clean.RData")

# OTU table with rare taxa filtered
save(otu_table_rare, file = "RData/16S/otu_rare_clean.RData")
```
