---
title: "Scarce features"
format: html
editor: visual
bibliography: references.bib
---

Scarce OTUs, defined by low prevalence across samples rather than low abundance, contribute substantially to data sparsity and are frequently driven by technical artefacts; filtering them using simple prevalence thresholds reduces noise and improves statistical robustness without targeting rare but consistently present taxa [@cao2021].

In R, low-prevalence OTUs are typically removed by calculating the proportion or number of samples in which each OTU is detected and pruning taxa that fall below a chosen prevalence threshold with `phyloseq::prune_taxa()`.

This section documents how to filter out scarce OTUs in `phyloseq` objects. To follow the example presented on this page directly on your computer, download the script `004_global_filter.R` and the example raw data by clicking the buttons below. Alternatively, you may copy the codes below into your own script, and use your own data.

```{r}
#| echo: false
#| output: false
library(downloadthis)
```

```{r}
#| layout: [[-10,35,-5,40,-10]]
#| echo: false
download_file(
  path = "../../scripts/004_global_filter.R",
  output_name = "004_global_filter",
  button_label = "Download script",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

download_file(
  path = "../../RData/16S/otu_ctrl_pos_clean.RData",
  output_name = "otu_ctrl_pos_clean",
  button_label = "Download example RData",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

### Removing scarce OTUs
