---
title: "Dealing with controls"
format: html
editor: visual
bibliography: references.bib
---

During DNA extraction, library preparation and sequencing, technical blanks, negative controls and positive controls may be added to detect contamination, confirm successful amplification of target genes, detect false results, and overall, verify that the full workflow (from extraction to bioinformatics) behaves as expected [@lear2018]. Field blanks (e.g., empty tubes left opened at the site or extraction of sample storage buffer) may also be included to enable removal of site or sampling material contaminants *a posteriori* [@tedersoo2022].

This section documents how we analyse control samples and translate those results into practical actions in the data set. Using the R package `decontam`, the workflow enables identification and removal of contaminant DNA sequences based on patterns observed in blanks and negative controls [@davis2018]. In addition, positive controls can be used to evaluate whether the observed community composition matches the expected taxonomic profile [@tedersoo2022].

The two downloadable scripts below (`003_negative_control.R` and `003_positive_control.R`) implement a standard approach for dealing with negative and positive controls in a `phyloseq` object.

```{r}
#| echo: false
#| output: false
library(downloadthis)
```

```{r}
#| layout: [[-15,25,-15,25,-15]]
#| echo: false
download_file(
  path = "../scripts/003_negative_control.R",
  output_name = "003_negative_control",
  button_label = "Download script 003_negative_control",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

download_file(
  path = "../scripts/003_positive_control.R",
  output_name = "003_positive_control",
  button_label = "Download script 003_positive_control",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

For a step-by-step explanation of how each control type is handled, see the sections below:

-   [Dealing with negative controls (and blanks)](controls-negative.qmd)
-   [Dealing with positive controls](controls-positive.qmd)
