---
title: "Dealing with negative controls"
format: html
editor: visual
---

This section documents how to inspect negative controls (or blanks) and remove contaminant DNA sequences using the R package `decontam` [@davis2018].

To follow the example presented on this page directly on your computer, download the script `003_negative_control.R` and the example raw data by clicking the buttons below. Alternatively, you may copy the codes below into your own script, and use your own data.

```{r}
#| echo: false
#| output: false
library(downloadthis)
library(plotly)
library(phyloseq)
library(decontam)
library(dplyr)
load("../RData/16S/otu_table_raw.RData")
```

```{r}
#| layout: [[-10,35,-5,40,-10]]
#| echo: false
download_file(
  path = "../scripts/003_negative_control.R",
  output_name = "003_negative_control",
  button_label = "Download script",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

download_file(
  path = "../RData/16S/otu_table_raw.RData",
  output_name = "16S_otu_table_raw",
  button_label = "Download example RData",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

### Inspect samples

Before using the package `decontam` to remove contaminants, we must first inspect samples (and controls) to determine if there are any visible problems. We can do this by comparing library size between samples, visualizing the prevalence of OTUs in samples and controls and visualizing the relative abundance of potential contaminants in all samples.

```{r}
#| eval: false
# Load OTU table
load("RData/16S/otu_table_raw.RData")
```

#### Library size

The library size (a.k.a. the sequencing depth) of each sample can be predicted based on the type of samples you work with. For example, with soil samples, we expect abundant microbial communities, with hundreds if not thousands of sequences per sample. On the other hand, negative controls and blanks should have very small library sizes.

```{r}
#| output: false
# Summarize sequencing depth per sample
seq_depth <- cbind.data.frame(sample_data(otu_table_raw)) %>%
  mutate(LibrarySize = sample_sums(otu_table_raw)) %>%
  # Sort the data set by LirarySize and create an Index to match the sorted data
  arrange(LibrarySize) %>% mutate(Index = row_number())

# Visualize LibrarySize as a function of Index, 
# to compare library size between samples
ggplot(data = seq_depth, aes(x = Index, y = LibrarySize, color = sample_type)) +
  geom_point()
```

```{r}
#| echo: false
plot_ly(data = seq_depth, x = ~Index, y = ~LibrarySize, color = ~sample_type,
        type = 'scatter', mode = 'markers', colors = 'Dark2', 
        text = ~paste("Library Size: ", LibrarySize, 
                      '<br>Sample ID:', Soil_ID))
```

We can see that, compared to the rest of the samples, all controls have a small library size. Nevertheless, if you zoom in on the negative controls, you can see that all negative controls have a library size larger than zero. Therefore, further analyses are required to know if these OTUs are contaminants.

Furthermore, one sample has a very small library size compared to other samples (and compared to controls). This sample should be further inspected to determine if the microbial community in this sample resemble that of other samples (i.e., replicates). If not, it should be discarded.

Note that positive controls have a larger library size than the negative controls, showing that sequencing performed well in these samples. This can be further confirmed in the next section: [Dealing with positive controls](controls-positive.qmd).

##### Inspect sample with small library size

```{r}
#| output: false
# Identify sample with a small library size
subset(seq_depth, sample_type == "sample" & LibrarySize < 10000)

# Subset OTU table with other samples from the same group (replicates)
# For this example, we will use samples that have roughly the same age (51-52)
otu_table.Age50 <- prune_samples(
  between(sample_data(otu_table_raw)$Age, 51,52), otu_table_raw)
```

```{r}
#| echo: false
cbind.data.frame(otu_table.Age50@sam_data[,c(4,5,7:9,11)][6:10]) %>% `rownames<-`(seq(5))
```

To compare these samples: first, convert OTU table from count data to relative abundance; then, remove taxa with less than 1% abundance (for clarity of the plot); and finally, visualize the taxa barcharts of these samples.

```{r}
#| warning: false
# Convert OTU table from count data to relative abundance
otu_table.Age50_rel <- transform_sample_counts(otu_table.Age50, function(x) x/sum(x))

# Remove taxa with less than 1% abundance
otu_table.Age50_rel <- filter_taxa(otu_table.Age50_rel, function(x) sum(x) > .01, TRUE)

# Visualize taxa barcharts
plot_bar(otu_table.Age50_rel, fill = "phylum")
```

The sample `X29239` looks quite different compared to the other samples and should therefore be removed.

#### Prevalence of OTUs in negative controls

```{r}
# For the following analyses, remove positive controls from the data set
otu_table_raw.noPos <- prune_samples(
  sample_data(otu_table_raw)$sample_type != "positive control", otu_table_raw)
```

First, visualize the prevalence of OTUs in samples and controls:

```{r}
# Make phyloseq objects of presence-absence in controls and true samples
otu_table.pa <- transform_sample_counts(otu_table_raw.noPos, function(x) 1*(x>0))
otu_table.pa.neg <- prune_samples(
  sample_data(otu_table.pa)$sample_type == "negative control", otu_table.pa)
otu_table.pa.sam <- prune_samples(
  sample_data(otu_table.pa)$sample_type == "sample", otu_table.pa)

# Make data.frame of prevalence in controls and true samples
otu_table.pa <- data.frame(pa.sam = taxa_sums(otu_table.pa.sam),
                           pa.neg = taxa_sums(otu_table.pa.neg))

# Plot the prevalence of OTU in negative controls vs true samples
ggplot(data=otu_table.pa, aes(x=pa.neg, y=pa.sam)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")
```

Many OTUs are present in at least one negative control ('Prevalence (Negative Controls)' \>= 1).

We must find out if there was a contamination in our samples from the method, or if the negative samples have been contaminated by the other samples.

```{r}
# For further analyses, convert OTU table from count data to relative abundance
otu_table_rel <- transform_sample_counts(otu_table_raw.noPos, function(x) x/sum(x))

# Identify OTUs present in negative controls
otu_neg <- filter(otu_table.pa, pa.neg > 0) %>% rownames()

# Filter OTU table to keep only OTUs found in negative controls
otu_table_neg <- prune_taxa(otu_neg, otu_table_rel)

# Visualize the abundance of these OTUs in all samples
plot_bar(otu_table_neg, fill = "phylum") +
  facet_grid(~sample_type, scales = "free", space = "free")
```

A commonly recommended practice with negative controls is to remove any OTU found in negative controls from the rest of the data set (<https://doi.org/10.1111/nph.12923>). However, as you can see in this demonstration, the OTUs found in the negative controls are found in all samples and represent \~25% of the community in all samples. Therefore, it is most likely that the OTUs found in the negative controls are present due to a contamination from the rest of the samples and not the other way around.

Another way to confirm this, is to look at the count data instead of the relative abundance:

```{r}
# Filter OTU table to keep only OTUs found in negative controls
otu_table_neg <- prune_taxa(otu_neg, otu_table_raw.noPos)

# Visualize the abundance of these OTUs in all samples
plot_bar(otu_table_neg, fill = "phylum") +
  facet_grid(~sample_type, scales = "free", space = "free")
```

Here we can see that the abundance of these contaminants in the negative controls is very small compared to true samples.

To deal with these OTUs and determine if they are truly contaminants, we can confirm (or infirm) our observations using the package `decontam`.

### R package `decontam`

The R package `decontam` uses simple statistical tools to identify contaminant automatically [@davis2018]. Visit [their webpage](https://bioconductor.org/packages/devel/bioc/vignettes/decontam/inst/doc/decontam_intro.html) for more information.

In this demonstration, we use the function `isContaminant` with the prevalence method to find which OTUs are contaminants.

```{r}
# First, add a TRUE/FALSE column "is.neg" to the sample_data
# to identify negative controls
sample_data(otu_table_raw.noPos)$is.neg <-
  sample_data(otu_table_raw.noPos)$sample_type == "negative control"

# Then use the function isContaminant with the default threshold = 0.1
contam.prev <- isContaminant(otu_table_raw.noPos, method="prevalence",
                             neg="is.neg", threshold = 0.1)

table(contam.prev$contaminant)
```

The method found 21 contaminant sequences at a threshold of 0.1 (default).

Increase the threshold to try and find more contaminants. This more aggressive threshold will identify as contaminants all sequences that are are more prevalent in negative controls than in true samples.

```{r}
# Use the function isContaminant with the more aggressive threshold = 0.5
contam.prev <- isContaminant(otu_table_raw.noPos, method="prevalence",
                             neg="is.neg", threshold = 0.5)

table(contam.prev$contaminant)
```

The method found 71 contaminant sequences at threshold = 0.5.

#### Inspect contaminants

Visualize the prevalence of these potential contaminants in samples and controls using the presence-absence table created earlier.

```{r}
# Add a TRUE/FALSE "contaminant" column to identify contaminants
otu_table.pa <- mutate(otu_table.pa, contaminant = contam.prev$contaminant)

# Plot the prevalence of OTU in negative controls vs true samples
ggplot(data=otu_table.pa, aes(x=pa.neg, y=pa.sam, color=contaminant)) + 
  geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")
```

The method identified OTUs that were mostly present in negative controls and few true samples (\<20%).

```{r}
# Filter OTU table to keep only OTUs suspected of being contaminants
contaminants <- filter(contam.prev, contaminant == TRUE) %>% rownames()
otu_table_contam <- prune_taxa(contaminants, otu_table_rel)

# Visualize the abundance of these OTUs in all samples
plot_bar(otu_table_contam, fill = "OTU") +
  facet_grid(~sample_type, scales = "free", space = "free") +
  theme(legend.position = "none")
```

After inspection of these OTUs, it is safe to consider them as contaminants and remove them from the data set. The other OTUs present in the negative controls were most likely coming from true samples and are therefore not considered contaminants.

### Clean and save OTU table

::: callout-tip
## remove_taxa

Homemade function `remove_taxa`: a simplified way of removing a selection of OTU from a `phyloseq` object.

```{r}
#| eval: false
remove_taxa <- function(badTaxa, physeq){
  allTaxa <- taxa_names(physeq)
  cleanTaxa <- allTaxa[!(allTaxa %in% badTaxa)]
  return(prune_taxa(cleanTaxa, physeq))
}
```
:::

```{r}
#| eval: false
# Remove samples with low sequencing depth (sampleID = X29239)
otu_table_raw <- prune_samples(
  sample_names(otu_table_raw) != "X29239", otu_table_raw)

# Remove contaminants (note that remove_taxa is a homemade function)
otu_ctrl_neg_clean <- remove_taxa(contaminants, otu_table_raw)

# Remove negative controls (and blanks)
otu_ctrl_neg_clean <- prune_samples(
  sample_data(otu_ctrl_neg_clean)$sample_type != "negative control", otu_ctrl_neg_clean)

# Save cleaned OTU table
save(otu_ctrl_neg_clean, file = "RData/16S/otu_ctrl_neg_clean.RData")
```
