---
title: "Filtering OTU tables"
format: html
editor: visual
bibliography: references.bib
---

This section presents two important ways of cleaning the OTU tables:

-   First, by going through a global filtering to remove singletons, rare or scarce OTUs, which are most often artifacts [@sgarbi2020; @nikodemova2023; @brown2015]. These also usually coincide with the OTUs with lowest BLAST accuracy [@tedersoo2022].
-   Then, by going through a taxonomy filtering, based on BLAST quality thresholds.

The two downloadable scripts below (`004_global_filter.R` and `005_taxonomy_filter.R`) implement a standard approach for cleaning OTU tables based on abundance and prevalence, or based on taxonomy and BLAST quality, in a `phyloseq` object.

```{r}
#| echo: false
#| output: false
library(downloadthis)
```

```{r}
#| layout: [[-10,30,-15,30,-10]]
#| echo: false
download_file(
  path = "../scripts/004_global_filter.R",
  output_name = "004_global_filter",
  button_label = "Download script 004_global_filter",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

download_file(
  path = "../scripts/005_taxonomy_filter.R",
  output_name = "005_taxonomy_filter",
  button_label = "Download script 005_taxonomy_filter",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

For a step-by-step explanation of each filtering procedure, see the sections below:

## Global filtering

A global filtering - based on abundance of OTUs across all samples (singletons and rare features), or based on the prevalence of OTUs across samples (scarcity) - is typically applied early to remove very low-count and low-prevalence features that are disproportionately driven by sequencing/PCR error and index misassignment, thereby reducing noise and improving downstream compositional and multivariate analyses.

### [Singletons and rare OTUs](filter/singletons-rare.qmd)

Singleton OTUs have been shown to be overwhelmingly composed of artefacts arising from PCR and sequencing errors, including chimeric molecules that are typically low-abundance and sample-restricted [@tedersoo2021; @sze2019]. Their removal has repeatedly been shown to have negligible effects on hypothesis-driven community comparisons while substantially reducing noise [@edgar2016; @nikodemova2023; @sgarbi2020]. On the other hand, retaining them is justifiable in studies explicitly targeting rare biosphere diversity, but requires careful manual curation [@brown2015].

Rare OTUs (beyond singletons) can hardly be distinguished from artefacts, thus, low-abundance OTUs are often excluded from the analysis. However, filtering approaches vary among studies, and there is no agreed-upon consensus as to what filtering thresholds should be employed [@yu2017; @poos2012; @nikodemova2023]. Nevertheless, rare OTUs, just like singletons, often represent residual artefacts, and removing them has minimal influence on community comparisons. Thus, filtering reduces the complexity of microbiome data while preserving their integrity in downstream analysis [@edgar2016; @nikodemova2023; @sgarbi2020; @cao2021]. Although rare taxa may in some contexts carry information on beta diversity, their retention is mainly justified in studies explicitly focused on rare biosphere ecology, where careful validation is required [@brown2015].

Rare OTUs are frequently removed using absolute abundance thresholds (e.g. \<5 or \<10 reads) [@edgar2016; @brown2015], or based on relative abundance thresholds (e.g. \<0.01%) [@nikodemova2023], while others filter rare OTUs based on statistical methods [@smirnova2018].

### [Scarce OTUs](filter/scarce.qmd)

Scarce (low-prevalence) OTUs are features detected in only a small fraction of samples (\<5%). The extreme sparsity of microbiome data sets poses a major analytical challenge and is largely attributable to sequencing artefacts, contamination, and/or sequencing errors [@cao2021]. Thus, scarce OTUs are commonly filtered using simple rule-of-thumb criteria (e.g. presence in at least k samples or k% of samples), which reduces noise and improves statistical stability [@cao2021]. This criterion does not relate to the relative abundance of the OTUs, and as such, rare OTUs may be prevalent microbial taxa.

In certain studies, higher thresholds for prevalence (\>25% of samples) may be used to define, for instance, the core microbiome [@simonin2020].

## Taxonomy filtering

\[INSERT HERE A GENERAL INTRODUCTION TO FILTERING BASED ON TAXONOMY AND BLAST QUALITY RESULTS\]

Filtering based on taxonomy and BLAST quality is applied to remove OTUs with ambiguous or low-confidence taxonomic assignments—typically indicated by low sequence similarity, poor alignment coverage, or inconsistent rank resolution—because such features are frequently artefactual or biologically uninterpretable and can bias ecological inference if retained (Altschul et al., 1990, DOI: 10.1016/S0022-2836(05)80360-2; Tedersoo et al., 2018, DOI: 10.1111/1462-2920.14487; Nilsson et al., 2019, DOI: 10.1111/1755-0998.13070).

-   [Bacteria & Archaea (16S)](filter/16S.qmd)
-   [Fungi (ITS)](filter/ITS.qmd)
-   [Soil animals (CO1)](filter/CO1.qmd)
